Miniproject - Combinatorics

### 2021-12-18 Analytic number theory/submission.tex
Added P_k,d proof to combinatorics

    More concretely, consider
    \begin{equation*}
        f: \N \to \C, \quad n \mapsto \begin{cases}
            1 & \text{if $n \equiv a \pmod q$} \\
            0 & \text{otherwise}
        \end{cases}
    \end{equation*}
    Note that this function is not multiplicative.
    After applying the first step, we are interested in the poles of the meromorphic function
    \begin{equation*}
        \sum_{n \geq 1} f(n) \Lambda(n) n^{-s}
    \end{equation*}
    I will not prove this here, but via partial summation, we can find a meromorphic continuation to $\Re(s) > 0$.

    However, here we get stuck.
    It might seem natural to consider the function
    \begin{equation*}
        \zeta_a(s) := \sum_{n \geq 1} f(n) n^{-s}
    \end{equation*}
    which also has a meromorphic continuation to $\Re(s) > 0$.
    Its derivative is clearly given by
    \begin{equation*}
        \zeta_a'(s) = \sum_{n \geq 1} f(n) \log(n) n^{-s}
    \end{equation*}
    To continue, we might try to consider
    \begin{equation*}
        \zeta_a'(s)\Bigl( \sum_{n \geq 1} \mu(n) f(n) n^{-s} \Bigr) = \sum_{n \geq 1} \frac {(f\log * f\mu)(n)} {n^s}
    \end{equation*}
    But since $f$ has no nice compatibility with multiplication or $*$, it is not easy to make something of this.
    In particular
    \begin{equation*}
        f\Lambda \neq f\log * f\mu \quad \text{and} \quad \frac 1 {\zeta_a(s)} \neq \sum_{n \geq 1} f(n) \mu(n) n^{-s}
    \end{equation*}
    Note that we find $n \equiv a \pmod q$ such that $n$ has no nontrivial factor $\equiv a \pmod q$.
    E.g. take $n = 9 \equiv 4 \pmod 5$, then
    \begin{equation*}
        \log(3) = (f\Lambda)(6) \neq \sum_{cd = 9} f(c)\log(c)f(d)\mu(d) = 0
    \end{equation*}
    Furthermore, the equality
    \begin{equation*}
        \frac 1 {\zeta_a(s)} = \sum_{n \geq 1} f(n) \mu(n) n^{-s}
    \end{equation*}
    would imply
    \begin{equation*}
        1 = \zeta_a(s) \sum_{n \geq 1} f(n) \mu(n) n^{-s} = \sum_{n \geq 1} \frac {(f * f\mu)(n)} {n^s}
    \end{equation*}
    which in turns implies that $f * f\mu = \delta$.
    However, we clearly find counterexamples for that, e.g. for $q = 5, a = 4$ find $(f * f\mu)(1) = 0$.
    The underlying reason is that the divisors of $n$ may lie in a different congruence class modulo $q$ than $n$.

>>>>

    More concretely, consider
    \begin{equation*}
        f: \N \to \C, \quad n \mapsto \begin{cases}
            1 & \text{if $n \equiv a \pmod q$} \\
            0 & \text{otherwise}
        \end{cases}
    \end{equation*}
    Note that this function is not multiplicative.
    After applying the first step, we are interested in the poles of the meromorphic function
    \begin{equation*}
        \sum_{n \geq 1} f(n) \Lambda(n) n^{-s}
    \end{equation*}
    However, here we get stuck.
    It might seem natural to consider the function
    \begin{equation*}
        \zeta_a(s) := \sum_{n \geq 1} f(n) n^{-s}
    \end{equation*}
    which has a meromorphic continuation to $\Re(s) > 0$ (I will not prove this here).
    Its derivative is clearly given by
    \begin{equation*}
        \zeta_a'(s) = \sum_{n \geq 1} f(n) \log(n) n^{-s}
    \end{equation*}
    To continue, we might try to consider
    \begin{equation*}
        \zeta_a'(s)\Bigl( \sum_{n \geq 1} \mu(n) f(n) n^{-s} \Bigr) = \sum_{n \geq 1} \frac {(f\log * f\mu)(n)} {n^s}
    \end{equation*}
    But since $f$ has no nice compatibility with multiplication or $*$, it is not easy to make something of this.
    In particular
    \begin{equation*}
        f\Lambda \neq f\log * f\mu \quad \text{and} \quad \frac 1 {\zeta_a(s)} \neq \sum_{n \geq 1} f(n) \mu(n) n^{-s}
    \end{equation*}
    Note that we find $n \equiv a \pmod q$ such that $n$ has no nontrivial factor $\equiv a \pmod q$.
    E.g. take $n = 9 \equiv 4 \pmod 5$, then
    \begin{equation*}
        \log(3) = (f\Lambda)(6) \neq \sum_{cd = 9} f(c)\log(c)f(d)\mu(d) = 0
    \end{equation*}
    Furthermore, the equality
    \begin{equation*}
        \frac 1 {\zeta_a(s)} = \sum_{n \geq 1} f(n) \mu(n) n^{-s}
    \end{equation*}
    would imply
    \begin{equation*}
        1 = \zeta_a(s) \sum_{n \geq 1} f(n) \mu(n) n^{-s} = \sum_{n \geq 1} \frac {(f * f\mu)(n)} {n^s}
    \end{equation*}
    which in turns implies that $f * f\mu = \delta$.
    However, we clearly find counterexamples for that, e.g. for $q = 5, a = 4$ find $(f * f\mu)(1) = 0$.
    The underlying reason is that the divisors of $n$ may lie in a different congruence class modulo $q$ than $n$.

### 2021-12-18 Combinatorics/submission.tex
Added P_k,d proof to combinatorics

After quite a long time trying to prove that this holds generally, I finally gave up and searched the literature.
There I found that $\mathcal{P}_{k, d}$ is also called divisor lattice and indeed we have the following statement.
\begin{proposition}
    $\mathcal{P}_{k, d}$ satisfies the conditions of Prop.~\ref{prop:characterization_lym}, i.e. an LYM-inequality.
\end{proposition}
\begin{proof}
    See e.g. \cite[Theorem 4.2.3]{anderson}.
\end{proof}
Interestingly, I have also found a variant of the LYM inequality (``pseudo-LYM inequality'') for the poset $\Pi_m$ that we studied earlier.
\begin{proposition}
    Let $\mathcal{A} \subseteq \Pi_m$ be a nonempty antichain. Then define
    \begin{equation*}
        a(n_1, ..., n_k) := |\bigl\{ A \in \mathcal{A} \ \bigm| \ |A| = k, \ |\{ S \in A \ | \ |S| = j \}| = n_j \bigr\}|
    \end{equation*}
    the number of partitions $A \in \mathcal{A}$ such that the classes in $A$ have sizes $n_1, ..., n_k$.
    Then
    \begin{equation*}
        \sum_{k \geq 1} \sum_{n_1, ..., n_k} \frac {a(n_1, ..., n_k) k!} {{m \choose n_1, ..., n_k} {m - 1 \choose k - 1}} \leq 1
    \end{equation*}
\end{proposition}
\begin{proof}
    See \cite{lym_partitions}.
\end{proof}
Note that the total number of partitions $\pi$ such that the classes in $\pi$ have sizes $n_1, ..., n_k$ is given by an expression involving the multinomial coefficient, namely
\begin{equation*}
    |\bigl\{ \pi \in \Pi_m \ \bigm| \ |\pi| = k, |\{ S \in \pi \ | \ |S| = j\}| = n_j \bigr\}| = {m \choose n_1, ..., n_k} \underbrace{ \prod_i \frac 1 {|\{ j \ | \ n_j = i\}|!} }_{=: \gamma(n_1, ..., n_k)}
\end{equation*}
Intuitively, the reason is that there are $m!$ ways to order the elements $[m]$, and after splitting such a sequence in chunks of sizes $n_1, ..., n_k$, there are $n_1! \cdot ... \cdot n_k!$ ways to order the element within the chunks.
Hence the multinomial coefficient
\begin{equation*}
    {m \choose n_1, ..., n_k}
\end{equation*}
gives the number of ways to distribute the elements of $[m]$ into ordered buckets of sizes $n_1, ..., n_k$.
A partition corresponds to unordered buckets, i.e. swapping chunks of equal size does not change the partition.
There are $1/\gamma(n_1, ..., n_k)$ ways to do so, and hence we get the expression mentioned above.

@@

Now note also that
\begin{equation*}
    {m - 1 \choose k - 1}
\end{equation*}
is exactly the number of different sequences $n_1 \leq ... \leq n_k$ by the ``stars and bars'' method.
Then one can see that the expression
\begin{equation*}    
    \sum_{k \geq 1} \sum_{n_1, ..., n_k} \frac {a(n_1, ..., n_k) k!} {{m \choose n_1, ..., n_k} {m - 1 \choose k - 1}} = \sum_{k \geq 1} \frac 1 {{m - 1 \choose k - 1}} \sum_{n_1, ..., n_k} \frac {a(n_1, ..., n_k)} {\frac 1 {k!} {m \choose n_1, ..., n_k}}
\end{equation*}
is very similar to the LYM inequality.
The only difference is that elements elements corresponding to different class sizes $n_1, ..., n_k$ might receive slightly different weight, depending on how many duplicate $n_i$ occur in that sequence.

>>>>

After quite a long time trying to prove that this holds generally, I finally gave up and searched the literature.
There I found that $\mathcal{P}_{k, d}$ is also called divisor lattice and indeed we have the following statement.
\begin{proposition}
    $\mathcal{P}_{k, d}$ satisfies the conditions of Prop.~\ref{prop:characterization_lym}, i.e. an LYM-inequality.
\end{proposition}
\begin{proof}
    The proof is similar to \cite[Theorem 4.2.3]{anderson}.
    However, I really like the proof and so we present it here in a somewhat modified form.

    We show condition (2) of Prop.~\ref{prop:characterization_lym}, so consider some pair of consecutive layers $L_{i - 1}, L_i$.
    Let further $F \subseteq L_i$. 
    We want to show that
    \begin{equation*}
        |L_{i - 1} \times F| \leq |L_i \times \partial F|
    \end{equation*}
    Let now $P' := \{ 0, ..., 2k \}^d$ and note that we have the decompositions
    \begin{align*}
        L_{i - 1} \times F &= \bigcup_{M \in P', r(M) = 2i - 1} \{ (u, a) \in L_{i - 1} \times F \ | \ u + a = M \} \quad \text{and} \\
        L_i \times \partial F &= \bigcup_{M \in P', r(M) = 2i - 1} \{ (v, b) \in L_i \times \partial F \ | \ v + b = M \}
    \end{align*}
    Hence it suffices to show that
    \begin{equation*}
        |\{ (u, a) \in L_{i - 1} \times F \ | \ u + a = M \}| \leq |\{ (v, b) \in L_i \times \partial F \ | \ v + b = M \}|
    \end{equation*}
    for each $M \in P', r(M) = 2i - 1$.

    To do so, we show the following, stronger claim by induction.
\end{proof}
\begin{lemma}
    Let $k_1, ..., k_d > 0$ and consider the poset
    \begin{equation*}
        P = \bigtimes_{i = 1}^d \{ 0, ..., k_d \}
    \end{equation*}
    Let further $M = (M_1, ..., M_d)$ with $M_i \leq 2k_i$ of rank $\sum_i M_i = 2l - 1$ and $F \subseteq L_l$ such that $M$ is an upper bound for $F$.
    Then
    \begin{equation*}
        |\{ (u, a) \in L_{l - 1} \times F \ | \ u + a = M \}| \leq |\{ (v, b) \in L_l \times \partial F \ | \ v + b = M \}|
    \end{equation*}
\end{lemma}
\begin{proof}
    Use induction on $\sum_i k_i$. 
    The base case is trivial, so assume $\sum_i k_i > 1$.

    \paragraph{Case 1} If there exists $j$ such that $M_j < k_j$ then clearly
    \begin{equation*}
        F, \partial F \in P' := \{ 0, ..., k_1 \} \times ... \times \{ 0, ..., k_j - 1 \} \times ... \times \{ 0, ..., k_d \}
    \end{equation*}
    and also $u \in P'$ for all $u \in L_{l - 1}$ with $u + a = M$ for some $a \in F$.
    Hence the induction hypothesis yields
    \begin{equation*}
        |\{ (u, a) \in L_{l - 1} \times F \ | \ u + a = M \}| \leq |\{ (v, b) \in L_l \times \partial F \ | \ v + b = M \}|
    \end{equation*}

    \paragraph{Case 2} If there exists $j$ such that $M_j > k_j \geq 1$, then consider
    \begin{align*}
        F' := &\{ a - e_j \ | \ a \in F, a(j) \neq 0 \} \in \\
        &P' := \{ 0, ..., k_1 \} \times ... \times \{ 0, ..., k_j - 1 \} \times ... \times \{ 0, ..., k_d \}
    \end{align*}
    Clearly $M - 2e_j \geq (k_1, ..., k_j - 1, ..., k_d) \geq a - e_j$ for all $a \in F$, and so $M - 2e_j$ is an upper bound for $F'$.
    By induction hypothesis, we now get that
    \begin{equation*}
        |\{ (u, a) \in L_{l - 2} \times F' \ | \ u + a = M - 2e_j \}| \leq |\{ (v, b) \in L_{l - 1} \times \partial F' \ | \ v + b = M - 2e_j \}|
    \end{equation*}
    However, observe that the map
    \begin{align*}
        \{ (u, a) \in L_{l - 2} \times F' \ | \ u + a = M - 2e_j \} &\to \{ (u, a) \in L_{l - 1} \times F \ | \ u + a = M \} \\
        (u, a) &\mapsto (u + e_j, a + e_j)
    \end{align*}
    is a bijection.
    The only nontrivial part is to see that it is surjective, so consider some $(u, a) \in L_{l - 1} \times F$ with $u + a = M$.
    Then $u(j), a(j) \leq k_j < M(j)$ and so $a(j) = M(j) - u(j) \geq 1$, hence $a - e_j \in F'$.
    Clearly $u - e_j \in L_{l - 2}$ and $u - e_j + a - e_j = M - 2e_j$.
    So $(u, a)$ has the preimage $(u - e_j, a - e_j)$.

    Similarly, see that the map
    \begin{align*}
        \{ (u, a) \in L_{l - 1} \times \partial F' \ | \ u + a = M - 2e_j \} &\to \{ (u, a) \in L_l \times \partial F \ | \ u + a = M \} \\
        (u, a) &\mapsto (u + e_j, a + e_j)
    \end{align*}
    is a bijection.
    Thus
    \begin{equation*}
        |\{ (u, a) \in L_{l - 1} \times F \ | \ u + a = M \}| \leq |\{ (v, b) \in L_l \times \partial F \ | \ v + b = M \}|
    \end{equation*}

    \paragraph{Case 3} If $M = (k_1, ..., k_d)$, see that $2l - 1 = \sum_i k_i$ and so the layers $L_{l - 1}, L_l$ must be the two unique middle layers.
    Now note that $|L_l| = |L_{l - 1}|$ and so it suffices to show that $|\partial F| \geq |F|$.
    However, this follows directly by the symmetric chain decomposition of the poset $P$.
    We have proven the symmetric chain decomposition only for $k_1 = ... = k_d$, but the general version can be found in \cite{anderson}.
\end{proof}
There are two parts I find interesting in this proof.
\begin{itemize}
    \item First of all, the idea to consider all $\bigtimes_i \{ 0, ..., k_i \}$ with possibly distinct $k_i$ makes it easy to use induction, as it suffices to decrease just one coordinate to use the induction hypothesis.
    \item The decomposition
    \begin{equation*}
        L_{i - 1} \times F = \bigcup_{M \in P', r(M) = 2i - 1} \{ (u, a) \in L_{i - 1} \times F \ | \ u + a = M \}
    \end{equation*}
    is just very clever, as it really transforms the set of tuples $(u, a)$ into multiple sets that are more or less $\{ a \in F \ | \ a \leq M \}$ and much easier to handle.
\end{itemize}
Interestingly, I have also found a variant of the LYM inequality (``pseudo-LYM inequality'') for the poset $\Pi_m$ that we studied earlier.
\begin{proposition}
    Let $\mathcal{A} \subseteq \Pi_m$ be a nonempty antichain. Then define
    \begin{equation*}
        a(n_1, ..., n_k) := |\bigl\{ A \in \mathcal{A} \ \bigm| \ |A| = k, \ |\{ S \in A \ | \ |S| = j \}| = n_j \bigr\}|
    \end{equation*}
    the number of partitions $A \in \mathcal{A}$ such that the classes in $A$ have sizes $n_1, ..., n_k$.
    Then
    \begin{equation*}
        \sum_{k \geq 1} \sum_{n_1, ..., n_k} \frac {a(n_1, ..., n_k) k!} {{m \choose n_1, ..., n_k} {m - 1 \choose k - 1}} \leq 1
    \end{equation*}
\end{proposition}
\begin{proof}
    See \cite{lym_partitions}.
\end{proof}
Note that the total number of partitions $\pi$ such that the classes in $\pi$ have sizes $n_1, ..., n_k$ is given by an expression involving the multinomial coefficient, namely
\begin{equation*}
    |\bigl\{ \pi \in \Pi_m \ \bigm| \ |\pi| = k, |\{ S \in \pi \ | \ |S| = j\}| = n_j \bigr\}| = {m \choose n_1, ..., n_k} \underbrace{ \prod_i \frac 1 {|\{ j \ | \ n_j = i\}|!} }_{=: \gamma(n_1, ..., n_k)}
\end{equation*}
Intuitively, the reason is that there are $m!$ ways to order the elements $[m]$, and after splitting such a sequence in chunks of sizes $n_1, ..., n_k$, there are $n_1! \cdot ... \cdot n_k!$ ways to order the element within the chunks.
Hence the multinomial coefficient
\begin{equation*}
    {m \choose n_1, ..., n_k}
\end{equation*}
gives the number of ways to distribute the elements of $[m]$ into ordered buckets of sizes $n_1, ..., n_k$.
A partition corresponds to unordered buckets, i.e. swapping chunks of equal size does not change the partition.
There are $1/\gamma(n_1, ..., n_k)$ ways to do so, and hence we get the expression mentioned above.

@@

Now note also that
\begin{equation*}
    {m - 1 \choose k - 1}
\end{equation*}
is exactly the number of different sequences $n_1 \leq ... \leq n_k$ summing to $m$ by the ``stars and bars'' method.
Then one can see that the expression
\begin{equation*}    
    \sum_{k \geq 1} \sum_{n_1, ..., n_k} \frac {a(n_1, ..., n_k) k!} {{m \choose n_1, ..., n_k} {m - 1 \choose k - 1}} = \sum_{k \geq 1} \frac 1 {{m - 1 \choose k - 1}} \sum_{n_1, ..., n_k} \frac {a(n_1, ..., n_k)} {\frac 1 {k!} {m \choose n_1, ..., n_k}}
\end{equation*}
is very similar to the LYM inequality.
The only difference is that elements elements corresponding to different class sizes $n_1, ..., n_k$ might receive slightly different weight, depending on how many duplicate $n_i$ occur in that sequence.
